Relationships in the extremes and their influence on competition and coexistence: Supporting information

### 2022-01-25 paper/SupMat.Rnw
edits to G4 methods paragraphs complete for now, over to Jasmin to review and compleete correspind SI sections
one more item needs to be added to the new sup mat section

%***DAN: Jasmin mentioned she had some math proving this, it needs to be added here.

@@

\section{Decomposing storage effect in empirical applications}\label{sect:Empircal_app}

The decomposition of storage effects and quantification of contributions of ATAs depends on 
the removal of asymmetric tail associations from the relationship between
coexistence drivers while maintaining their correlation and marginal distributions. 
In the paper section \ref{MT-section:theory}, we describe analytical methods for generating the partial sharp variables, $E_i^\nat$ and $C_{i \bs i}^\nat$, which are distributed in the same way as the defined $E_i$ and $C_{i \bs i}$, but are symmetric in the tails and have a normal copula structure.
The protocol for samples of coexistence drivers in empirical applications is as follows.
Let $\olE_i$ be a sample of a environmental fluctuation and $\olC_{i \bs i}$ be a sample of competitive pressure. Let $\Theta_{\olE}$ and $\Theta_{\olC}$ be standard normal distributions obtained by normalizing the ranks of $\olE_i$ and $\olC_{i \bs i}$, 
such that $\Theta_{\olE} = \frac{rank(\olE_i)}{n_{\olE_i}+1} \circ \varphi^{-1}$ and $\Theta_{\olC} = \frac{rank(\olC_{i \bs i})}{n_{\olC_{i \bs i}}+1} \circ \varphi^{-1}$, where $\varphi$ is the CDF of a standard normal distribution. The normalized correlation of $\olE_i$ and $\olC_{i \bs i}$, denoted by $r$, is the covariance of $\Theta_{\olE}$ and $\Theta_{\olC}$.  
We then simulate an $S$ number of bivariate normal distributions $N_s\left(\vec{0}, \Sigma \right)$
with covariance matrix $\Sigma = \begin{pmatrix} 1&\ddot{\rho} \\ \ddot{\rho}&1\end{pmatrix}$. 
The columns of $(\olE_{i}, \olC_{i \bs i})$ are permutated and saved $S$ amount of times, 
such that the rank of each column matches with the rank of each column of $N_s$. 
To obtain $(\olE_{i}^\nat, \olC_{i \bs i}^\nat)$, the median of the permuations is computed. 
Here, $\ddot\rho$ can be chosen a number of ways such that the Pearson, Spearman, or Kendall 
correlation of $(\olE_{i}^\nat, \olC_{i \bs i}^\nat)$ matches that of $(\olE_{i}, \olC_{i \bs i})$
or such that $\ddot\rho = r$. The code for this protocol is in \url{"diatom/partialSharp_fxns.R"} and was used in our application to the diatom chemostat system.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig3_qij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig3} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig3} of the main text, but using 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3qij,5)}, so are not displayed. \label{fig:fig3_alt}}
\end{figure}
%***DAN: Jasmin, in the caption of Fig. 3 in the main text there is a statment about standard
%error. We need another such statement here, but for the standard errors that applied here.

>>>>

%***DAN: Jasmin mentioned she had some math proving this, it needs to be added here.
%It has been added but Dan needs to edit it.

@@

\section{Details of diatom model setup}\label{sect:diatom_model}

%***DAN: Jasmin, please start on this section when you get a chance. Here are some topics that need
%to be covered:

%1) Precise delineation of the functions Q_i, V_i, K_i which were used. There are different options considered
%all over the place in the code and by Ellner et al (2019) and by the original diatom study, so we have to
%be very clear here. To give one example of the need for care, here, Ellner et al (2019) state on p. 6
%of their paper that K_1 and K_2 were constant for the range of temperature considered, and they state
%that they took constant values for these things. However, we used a larger range of temperatures. 
%Depending on which of K1fun/K2fun versus K1flatfun/K2flatfun (from DDataAndVKQfuns.R) was actually used
%may make a difference, so we have to specify. Note that even K1flatfun/K2flatfun were not actually flat
%for the whole temperature range 6-24 degrees C, though they were for the range of temperature Ellner
%et al considered. We should not say we used flat K1 and K2 functions unless we are sure we only ever
%used temperatures in the flat ranges of these functions. Better just to describe the functions we
%actually used, where the best way to describe them *may* be by plotting them. Anyway, what you need to
%do is make sure you know what versions of each of these 6 functions your code actually used, and then
%describe the functions precisely in this section.

%2) You need to specify what choices were made for E and C. The text below, which I cut-and-pasted
%from the main text, does some of this. Please use an edited version of that text here, edited as you
%see fit if clarity needs to be improved and to fit the text in with the rest of this new section.
%You ought to add a bit more exlanation to the text, so one understand why the choice of C make sense
%and why r is what you've written there.

%3) You need to outline how the quantities E_i and C_i\i were obtained from simulations.

%***DAN: Some text and notes immediately below were cut and pasted from the main text, in case 
%having this material helps in writing this new SI section
%
%and here, we consider varying parameters of temperature fluctuations and further decompose the 
%coexistence of \emph{C. pseudostelligera} and \emph{F. crotonensis} to account for contributions 
%of ATAs and ATA effects as described above. Our decomposition is defined by $E(t)=\theta(t)$, 
%$C_j(t)=(K_j(t)+S(t))/S(t)$ and $r_j(E_j,C_j)=\frac{V_j(E)}{C_j} -D$. Thus in our exploration, 
%environmental fluctuation is temperature fluctuation. We first considered the scenario where 
%\emph{F. crotonensis} is the invader and emph{C. pseudostelligera} is the resident to compute 
%the coexistence decomposition for \emph{F. crotonensis}. 
%put reverse in sup mat?
%***DAN: Yes, I think that would be a good idea. I forgot about that - we were able to only consider 1 invading 2 for 
%the lottery model, because of some simplifying assumptions we were able to make, but cannot really do that here, right?
%***JASMIN: Right. Species in the lottery model only differed in mu but here they differ in they're VKQ functions.
%*** JASMIN: Note to to run the reverse invader-resident scenario and put in SUpMat!
%do we need to explain how we separated the asymmetries using the sims??
%***DAN: I am not sure what you mean by this question, Jasmin
%***JASMIN: I mean should we explain the methodology of the code we wrote that is implemented to compute the 
% ATA contribution. The functions alignranks, normcor...  (in partialSharp_fxns.R) 
%***DAN: Yes, we are going to have to do that

\section{Decomposing storage effects in the diatom case study}\label{sect:Empircal_app}

The decomposition of storage effects and quantification of contributions of ATAs depends on 
the removal of asymmetric tail associations from the bivariate distribution
$(E_i,C_{i \bs i})$ while maintaining its correlation and marginal distributions.
In the Theory section of the main text, we decribed methods for this removal process
when the distributions of $E_i$ and $C_{i \bs i})$ are explicitly known, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{}), 
we only have samples of $E_i$ and $C_{i \bs i})$

;
a modified distribution $E_i^\nat$ and $C_{i \bs i}^\nat$

In the paper section \ref{MT-section:theory}, we describe analytical methods for generating the partial sharp variables, $E_i^\nat$ and $C_{i \bs i}^\nat$, which are distributed in the same way as the defined $E_i$ and $C_{i \bs i}$, but are symmetric in the tails and have a normal copula structure.
The protocol for samples of coexistence drivers in empirical applications is as follows.
Let $\olE_i$ be a sample of a environmental fluctuation and $\olC_{i \bs i}$ be a sample of competitive pressure. Let $\Theta_{\olE}$ and $\Theta_{\olC}$ be standard normal distributions obtained by normalizing the ranks of $\olE_i$ and $\olC_{i \bs i}$, 
such that $\Theta_{\olE} = \frac{rank(\olE_i)}{n_{\olE_i}+1} \circ \varphi^{-1}$ and $\Theta_{\olC} = \frac{rank(\olC_{i \bs i})}{n_{\olC_{i \bs i}}+1} \circ \varphi^{-1}$, where $\varphi$ is the CDF of a standard normal distribution. The normalized correlation of $\olE_i$ and $\olC_{i \bs i}$, denoted by $r$, is the covariance of $\Theta_{\olE}$ and $\Theta_{\olC}$.  
We then simulate an $S$ number of bivariate normal distributions $N_s\left(\vec{0}, \Sigma \right)$
with covariance matrix $\Sigma = \begin{pmatrix} 1&\ddot{\rho} \\ \ddot{\rho}&1\end{pmatrix}$. 
The columns of $(\olE_{i}, \olC_{i \bs i})$ are permutated and saved $S$ amount of times, 
such that the rank of each column matches with the rank of each column of $N_s$. 
To obtain $(\olE_{i}^\nat, \olC_{i \bs i}^\nat)$, the median of the permuations is computed. 
Here, $\ddot\rho$ can be chosen a number of ways such that the Pearson, Spearman, or Kendall 
correlation of $(\olE_{i}^\nat, \olC_{i \bs i}^\nat)$ matches that of $(\olE_{i}, \olC_{i \bs i})$
or such that $\ddot\rho = r$. The code for this protocol is in \url{"diatom/partialSharp_fxns.R"} and was used in our application to the diatom chemostat system.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig3_qij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig3} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig3} of the main text, but using 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3qij,5)}, so are not displayed. \label{fig:fig3_alt}}
\end{figure}
